### 2024-07-05 parts/conclusion.tex
starving the beast

Let us conclude with a more general remark on the status of commutative input
in the study of unary output polyregular functions. Notice that semantic
characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
polynomial}) arise from pre-composition with a \kl{commutative}
\kl{star-free polyregular function}, as formally stated in
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}. Following
this idea, we conjecture that $\ZPoly$ and $\NPoly$ can be described as
combinations of commutative functions, leading to
\cref{z-poly-commutative-encoding:conj}, which would be a non-commutative
version of \cref{decompose-polynomial:lem}. For the subclass of 
\emph{polyblind functions} \cite{LENP21,DOUE22}, parts of the conjecture holds,
because they essentially behave as commutative functions \cite[Theorem
6.12]{DOUE23}.

@@

>>>>

Let us conclude with a more general remark on the status of commutative input
in the study of unary output polyregular functions. Notice that semantic
characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
polynomial}) arise from pre-composition with a \kl{commutative}
\kl{star-free polyregular function}, as formally stated in
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.

@@

Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
immediately, and $\NSF$ inside $\NPoly$ would immediately follow. On the one
hand, one can guess a candidate function and check for equivalence, on the
other hand, one can guess a \kl{commutative} \kl{star-free polyregular
function} and check membership (which is decidable thanks to this paper).

### 2024-07-05 parts/starfree.tex
starving the beast

\AP To actually state the lemma, we will need some notations that will
formalize the intuitive idea of counting tuples of integers modulo some
constant. Let $\omega \in \Nat$, we define the set $\intro*\ModuloTypes[\omega]
\defined \set{0, \dots, \omega^2} \times \set{0, \dots, \omega - 1}$ of
\intro{types modulo $\omega$}. To a number $n \in \Nat$, we associate its
\reintro{$\omega$-type} written $\intro*\moduloType[\omega](n)$ which is
defined as the pair $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the  remainder
of the Euclidean division of $n$ by $\omega$. This notion of type is lifted to
vectors in $\Nat^p$ for any $p \in \Nat$ by pointwise application.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
        = P_{\moduloType[\omega](X_1, \dots, X_n)}
        \left(
            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}

\textbf{TODO: prove the following conjecture to conclude on polynomials in general
and not just Z coefficients.}
\begin{conjecture}
    Let $\vec{X}$ be a vector of variables, and $P \in \Rel[\vec{X}]$ be a
    polynomial. The following properties are decidable
    \begin{itemize}
        \item Is $P$ \kl{represented} by a \kl{$\Rel$-polyregular function}?
        \item Is $P$ \kl{represented} by a \kl{$\Nat$-polyregular function}?
    \end{itemize}
\end{conjecture}

>>>>

\AP To actually state the lemma, we will need some notations that will
formalize the intuitive idea of identifying tuples of integers modulo some
constant. Let $\omega \in \Nat_{\geq 1}$, we define the set
$\intro*\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
\dots, \omega - 1}$ of \intro{types modulo $\omega$}. To a number $n \in \Nat$,
we associate its \reintro{$\omega$-type} written
$\intro*\moduloType[\omega](n)$ which is defined as the pair $(t, r)$ where $t
= \min (n, \omega^2)$ and $r$ is the  remainder of the Euclidean division of
$n$ by $\omega$. This notion of type is lifted to vectors in $\Nat^p$ for any
$p \in \Nat$ by pointwise application.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq \omega$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{zsf-npoly-nsf:thm:up:item} $f$ is \kl{ultimately polynomial},
        \item \label{zsf-npoly-nsf:thm:zsf:item} $f \in \ZSF$,
        \item \label{zsf-npoly-nsf:thm:nsf:item} $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}
\begin{proof}
    The implication \cref{zsf-npoly-nsf:thm:zsf:item} $\Rightarrow$
    \cref{zsf-npoly-nsf:thm:zsf:item} is immediate since $\NSF \subseteq \ZSF$.
    Furthermore,
    \cref{zsf-npoly-nsf:thm:zsf:item} implies \cref{zsf-npoly-nsf:thm:up:item}
    following previous results for \kl{star-free $\Rel$-polyregular functions}
    \cite{CDTL23}.

    For the implication \cref{zsf-npoly-nsf:thm:up:item} $\Rightarrow$
    \cref{zsf-npoly-nsf:thm:nsf:item}, let us assume that $f$ is \kl{ultimately
    polynomial}. We prove the result by induction on the size of the alphabet
    $\Sigma$. By definition, there exists $N_0 \in \Nat$, and $P \in
    \Rat[\seqof{X_a}{a \in \Sigma}]$ such that: \begin{equation*} f\left(
        \prod_{a \in \Sigma} a^{X_a} \right) = P(\seqof{X_a}{a \in \Sigma})
        \quad \text{ when } \forall a \in \Sigma, X_a \geq N_0 \quad .
    \end{equation*} It is clear that $P$ is \kl{represented} by a
    \kl{$\Nat$-polyregular function}, namely, $f$ partially applied to the word
    $\prod_{a \in \Sigma} a^{N_0}$, and therefore is in fact represented by a
    \kl{star-free $\Nat$-polyregular function} thanks to
    \cref{decide-rat-poly-npoly:cor}.
    For every letter $a \in \Sigma$ and number $0 \leq n \leq N_0$, there exists, by induction hypothesis,
    a \kl{star-free $\Nat$-polyregular function} $g_a^n$ that represents the
    function $f_{a^n} \colon (\Sigma \setminus \set{a})^* \to \Rel$
    that maps $u \in \Sigma \setminus \set{a}$ to $f(a^n u)$.

    Let us now define the function $g \colon \Sigma^* \to \Rel$ as follows:
    \begin{equation*}
        g \colon w \mapsto \begin{cases}
            g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
            P(\seqof{\card[a]{w}}{a \in \Sigma}) & \text{ otherwise }
        \end{cases}
    \end{equation*}
    We conclude that $g$ is a \kl{star-free $\Nat$-polyregular function}, and by definition
    $f = g$.
\end{proof}
